---
title: "Bootstrapping Censored Data: Explicit Implementations"
subtitle: "Transparent re-implementation of `censboot_lau.R`"
author: "censor-bootstrap project (original code: Jacqueline Rudolph)"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(survival)
library(here)
i_am("code/censboot_explicit.Rmd")
set.seed(123)
```

# Purpose

`code/censboot_lau.R` estimates the **10-year risk of AIDS or death** (censoring at
treatment) by baseline injection-drug use (`BASEIDU`) in the Lau (2009) WIHS data, and
gets standard errors four ways: a classic bootstrap, a case-based bootstrap, a
conditional bootstrap, and a "weird" bootstrap — all through `boot::boot` and
`boot::censboot`. 

Those functions are black boxes. Here we reproduce the same analysis
with more explicit user-defined functions to make the resampling steps visible. 

A validation appendix runs the original package code verbatim to confirm agreement.

The `classic` and `case` blocks of the original are the *same* algorithm (ordinary nonparametric
resampling) run through two different wrappers, so the three distinct schemes are
**ordinary**, **conditional**, and **weird**.

```{r data}
raw <- read.csv(here("data", "wihs_public_lau2009.csv"))
dat <- data.frame(
  t       = raw$t,
  delta   = as.numeric(raw$eventtype == 2),   # 1 = AIDS/death; 0 = censored or competing event
  BASEIDU = raw$BASEIDU
)
tau <- 10                                       # risk horizon, in years
R   <- 1000                                     # bootstrap replicates

head(dat)

```

# The estimand and the analytic reference

Every method targets the same statistic: for each `BASEIDU` stratum, the risk
$\theta_g = 1 - \hat S_g(\tau)$ at $\tau = 10$ years, where $\hat S_g$ is the Kaplan–Meier
survival curve in stratum $g$.

Isolate that in one function used everywhere:

```{r estimand}
# Risk at tau for each stratum (g = 0, 1), returned in order c(IDU=0, IDU=1)
risk_at_t <- function(data, tau = 10) {
  vapply(c(0, 1), function(g) {
    sf <- survfit(Surv(t, delta) ~ 1, data = data[data$BASEIDU == g, , drop = FALSE])
    1 - summary(sf, times = tau, extend = TRUE)$surv
  }, numeric(1))
}

# Greenwood: point risk and analytic SE (SE of 1 - S equals SE of S)
greenwood_ref <- function(data, tau = 10) {
  out <- sapply(c(0, 1), function(g) {
    sf <- survfit(Surv(t, delta) ~ 1, data = data[data$BASEIDU == g, , drop = FALSE])
    s  <- summary(sf, times = tau, extend = TRUE)
    c(risk = 1 - s$surv, se = s$std.err)
  })
  colnames(out) <- c("IDU=0", "IDU=1")
  out
}

gw <- greenwood_ref(dat, tau)
gw
```

Two small helpers used by the resampling schemes below: a safe categorical sampler
(avoids R's `sample()` "single value" trap, that selects a range from 1:value if 
only a single value is supplied), and a routine that reads the jump masses
off a Kaplan–Meier curve — i.e. that turns an estimated survival function into the
discrete distribution you simulate from.

```{r helpers}
# Draw n values from the support `times` with probabilities `prob`
rcat <- function(times, prob, n = 1) {
  times[sample.int(length(times), n, replace = TRUE, prob = prob)]
}

# Discrete distribution implied by a KM curve: P(T = t_j) = S(t_{j-1}) - S(t_j)
km_atoms <- function(time, status) {
  sf   <- survfit(Surv(time, status) ~ 1)
  jump <- c(1, head(sf$surv, -1)) - sf$surv     # 0 at times with no "event" of this type
  data.frame(time = sf$time, prob = jump)
}
```

# Ordinary (case) bootstrap — "Efron's method"

This is the classic nonparametric resampling bootstrap. Treat the $n$ observed 
pairs $(t_i, \delta_i)$ as the population and draw
a new sample of size $n$ *with replacement*, then recompute the risk. 

Efron (1981) showed this is equivalent to independently simulating a fresh event time from $\hat F$ and a fresh censoring time from $\hat G$ for each record and keeping the smaller — so both the event
times and the censoring pattern vary from replicate to replicate. 

It assumes nothing beyond independent, identically distributed subjects. 

This is what both the original `classic` (`boot`) and `case` (`censboot(sim = "ordinary")`) blocks do.

Pseudocode: 

```
for r in 1..R:
    draw indices i_1..i_n uniformly with replacement from 1..n
    D* <- data[i_1..i_n, ]
    theta*_r[g] <- 1 - KM_g(tau) fit on D*      for each stratum g
SE[g] <- sd_r( theta*_r[g] )
```

```{r ordinary}
boot_ordinary <- function(data, R = 1000, tau = 10) {
  n <- nrow(data)
  replicate(R, {
    idx <- sample.int(n, n, replace = TRUE)     # resample whole records
    risk_at_t(data[idx, , drop = FALSE], tau)
  })
}

set.seed(123)
ord_reps <- boot_ordinary(dat, R, tau)
ord_se   <- apply(ord_reps, 1, sd)
ord_se
```

# Conditional bootstrap

**How it works.** The conditional bootstrap holds the *censoring pattern* closer to what
was observed. For each subject it draws a brand-new event time from the Kaplan–Meier
event distribution $\hat F$ (within that subject's stratum), but treats censoring
conditionally: subjects observed to be censored keep their exact censoring time, while
subjects observed to fail are assigned a hypothetical censoring time drawn from the
censoring distribution $\hat G$ restricted to times *after* their observed follow-up. The
new observation is the minimum of the two. Because $\hat F$ may not reach 1 when the
largest time is censored, a probability atom beyond the last observed time absorbs the
residual mass (equivalent to Davison & Hinkley's "add a record past the maximum" device;
`censboot` does this internally — cf. the commented `bind_rows` line in the original).

```
Precompute  F_g <- KM of event times within stratum g       (status = delta)
            G   <- KM of censoring times, pooled             (status = 1 - delta)
for r in 1..R:
    for subject i in stratum g:
        Y0*_i <- draw from F_g                               # new latent event time
        if delta_i == 0:  C*_i <- t_i                        # keep observed censoring
        else:             C*_i <- draw from G on (t_i, inf)  # hypothetical censoring
        t*_i     <- min(Y0*_i, C*_i)
        delta*_i <- 1{ Y0*_i <= C*_i }
    theta*_r[g] <- 1 - KM_g(tau) fit on (t*, delta*)
SE[g] <- sd_r( theta*_r[g] )
```

```{r conditional}
boot_conditional <- function(data, R = 1000, tau = 10) {
  n <- nrow(data)
  # F-hat: event-time distribution within each stratum
  Fhat <- lapply(c(0, 1), function(g) {
    d <- data[data$BASEIDU == g, ]
    km_atoms(d$t, d$delta)
  })
  names(Fhat) <- c("0", "1")
  # G-hat: censoring-time distribution, pooled across strata
  Ghat      <- km_atoms(data$t, 1 - data$delta)
  Ghat      <- Ghat[Ghat$prob > 0, ]            # keep only the censoring atoms
  tail_time <- max(data$t) + 1                  # absorbs residual mass of F-hat

  one_rep <- function() {
    # Step 1: new event times from the stratum-specific F-hat
    y0 <- numeric(n)
    for (g in c(0, 1)) {
      idx <- which(data$BASEIDU == g)
      A   <- Fhat[[as.character(g)]]
      y0[idx] <- rcat(c(A$time, tail_time), c(A$prob, 1 - sum(A$prob)), length(idx))
    }
    # Step 2: censoring times, conditional on the observed pattern
    cstar <- data$t                             # delta == 0: keep observed censoring time
    for (i in which(data$delta == 1)) {         # delta == 1: draw C > observed t_i
      elig     <- Ghat[Ghat$time > data$t[i], ]
      cstar[i] <- if (nrow(elig) == 0) Inf else rcat(elig$time, elig$prob, 1)
    }
    # Step 3: combine and evaluate the estimand
    risk_at_t(data.frame(t = pmin(y0, cstar),
                         delta = as.numeric(y0 <= cstar),
                         BASEIDU = data$BASEIDU), tau)
  }
  replicate(R, one_rep())
}

set.seed(123)
cond_reps <- boot_conditional(dat, R, tau)
cond_se   <- apply(cond_reps, 1, sd)
cond_se
```

# Weird bootstrap

**How it works.** The weird bootstrap (Andersen, Borgan, Gill & Keiding 1993) works on the
counting-process representation and holds *everything* fixed except the number of events.
At each observed failure time $u_j$ the risk set $n_j$ and the observed number of events
$d_j$ are known; it redraws the events there as an independent
$\text{Binomial}(n_j, d_j/n_j)$ and rebuilds the survival curve as the product-limit
$\hat S^*(\tau) = \prod_{u_j \le \tau}(1 - d_j^*/n_j)$. Nothing about who is at risk or when
censoring happens changes — only the event counts fluctuate — so it reproduces the
Nelson–Aalen / Greenwood variance by construction. It is "weird" because a replicate
corresponds to no actual set of individuals.

```
Precompute per stratum g:  failure times u_1<...<u_m (<= tau),
                           risk sets n_j, event counts d_j
for r in 1..R:
    for each stratum g, each failure time u_j:
        d*_j <- Binomial(n_j, d_j / n_j)        # risk set n_j held fixed
    theta*_r[g] <- 1 - prod_j (1 - d*_j / n_j)
SE[g] <- sd_r( theta*_r[g] )
```

```{r weird}
boot_weird <- function(data, R = 1000, tau = 10) {
  # Per stratum, capture the fixed risk sets and event counts up to tau
  strata <- lapply(c(0, 1), function(g) {
    sf <- survfit(Surv(t, delta) ~ 1, data = data[data$BASEIDU == g, ])
    keep <- sf$n.event > 0 & sf$time <= tau
    list(n = sf$n.risk[keep], d = sf$n.event[keep])
  })
  replicate(R, vapply(strata, function(s) {
    dstar <- rbinom(length(s$d), size = s$n, prob = s$d / s$n)   # resample events only
    1 - prod(1 - dstar / s$n)                                    # 1 - S*(tau)
  }, numeric(1)))
}

set.seed(123)
weird_reps <- boot_weird(dat, R, tau)
weird_se   <- apply(weird_reps, 1, sd)
weird_se
```

# Results: all four standard errors side by side

```{r results}
comparison <- data.frame(
  Stratum          = c("IDU = 0", "IDU = 1"),
  `Risk at 10y`    = round(gw["risk", ], 3),
  `SE: Greenwood`  = round(gw["se", ],   3),
  `SE: Ordinary`   = round(ord_se,       3),
  `SE: Conditional`= round(cond_se,      3),
  `SE: Weird`      = round(weird_se,     3),
  check.names = FALSE, row.names = NULL
)
knitr::kable(comparison, caption = "10-year risk of AIDS/death and its standard error.")
```

The weird bootstrap should sit essentially on top of Greenwood; the ordinary and
conditional schemes approximate it up to Monte-Carlo error. Where the conditional and
ordinary SEs *diverge* from Greenwood is the substantive question the paper is after —
e.g. whether they carry variability (such as in estimated weights) that the analytic
formula and the weird scheme do not.

# Appendix: validation against the original package code

This chunk runs `censboot_lau.R` verbatim through `boot`/`censboot` and lines its standard
errors up against the explicit functions above. It is set with `error=TRUE` so a failure
here never blocks the knit. The ordinary and conditional schemes agree with their explicit
counterparts up to Monte-Carlo noise (independent random streams, so they will not match to
the last digit). The **weird** row is the exception worth noting: the packaged
`sim = "weird"` here comes out systematically *below* both Greenwood and the explicit weird
bootstrap — which, by construction, reproduces Greenwood — and the gap persists as $R$
grows, so it is a feature of how the original `surv.fun3`/`censboot` weird call is wired
rather than a seed artifact. Treat the explicit `boot_weird()` above as the reference.

```{r validation, error=TRUE}
library(boot)

# --- classic: boot() case resampling ---
surv.fun <- function(data, d) risk_at_t(data[d, ], tau)
set.seed(123); classic <- boot(dat, surv.fun, R = R)

# --- case / cond: censboot() ---
surv.fun2   <- function(data) risk_at_t(data, tau)
event.surv  <- survfit(Surv(t, delta) ~ BASEIDU, data = dat)
censor.surv <- survfit(Surv(t - 0.0001 * delta, 1 - delta) ~ 1, data = dat)
set.seed(123); case <- censboot(data.frame(dat), surv.fun2, R = R)
set.seed(123); cond <- censboot(data.frame(dat), surv.fun2, R = R, strata = dat$BASEIDU,
                                F.surv = event.surv, G.surv = censor.surv, sim = "cond")

# --- weird: censboot() on the (time, status) matrix, exactly as in the folder ---
surv.fun3 <- function(data, str) {
  surv <- survfit(Surv(data[, 1], data[, 2]) ~ str)
  res  <- data.frame(idu  = c(rep(0, surv$strata[[1]]), rep(1, surv$strata[[2]])),
                     time = surv$time, risk = 1 - surv$surv)
  res2 <- res[res$time <= tau, ]
  res2[!duplicated(res2$idu, fromLast = TRUE), "risk"]
}
set.seed(123); weird <- censboot(cbind(dat$t, dat$delta), surv.fun3, R = R,
                                 strata = dat$BASEIDU, F.surv = event.surv, sim = "weird")

validation <- data.frame(
  Stratum            = c("IDU = 0", "IDU = 1"),
  `boot (classic)`   = round(apply(classic$t, 2, sd), 3),
  `censboot ordinary`= round(apply(case$t,    2, sd), 3),
  `censboot cond`    = round(apply(cond$t,    2, sd), 3),
  `censboot weird`   = round(apply(weird$t,   2, sd), 3),
  check.names = FALSE, row.names = NULL
)
knitr::kable(validation, caption = "Package standard errors (compare to the explicit table above).")
```

---

*Conceptual background: the vault Main Note "Bootstrapping Censored Data". Full annotated
bibliography: `_corpus/censor-bootstrap_references.md`.*
