---
title: "Bootstrapping Censored Data: Explicit Implementations"
subtitle: "Running weighted survival"
author: "censor-bootstrap project (Jacqueline Rudolph, Ashley Naimi)"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    theme: readable
knit: (function(inputFile, encoding) { 
      out_dir <- 'doc';
      rmarkdown::render(inputFile,
                        encoding=encoding, 
                        output_file=file.path("..", out_dir, 'censboot_explicit_ipw.html')) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(survival)
library(here)
i_am("code/censboot_explicit.Rmd")
set.seed(123)
```

# Purpose

`code/censboot_lau.R` estimates the **10-year risk of AIDS or death** (censoring at
treatment) by baseline injection-drug use (`idu`) in the Lau (2009) WIHS data, and
gets standard errors four ways: a classic bootstrap, a case-based bootstrap, a
conditional bootstrap, and a "weird" bootstrap — all through `boot::boot` and
`boot::censboot`. 

Those functions are black boxes. Here we reproduce the same analysis
with more explicit user-defined functions to make the resampling steps visible. 

A validation appendix runs the original package code verbatim to confirm agreement.

The `classic` and `case` blocks of the original are the *same* algorithm (ordinary nonparametric
resampling) run through two different wrappers, so the three distinct schemes are
**ordinary**, **conditional**, and **weird**.

```{r data}
raw <- read.csv(here("data", "wihs_public_lau2009.csv"))
dat <- data.frame(
  t       = raw$t,
  delta   = as.numeric(raw$eventtype == 2),   # 1 = AIDS/death; 0 = censored or competing event
  idu = raw$BASEIDU,
  age = raw$ageatfda,
  black = raw$black,
  cd4nadir = raw$cd4nadir
)
tau <- 10                                       # risk horizon, in years
R   <- 1000                                     # bootstrap replicates

head(dat)

```

# The estimand and the analytic reference

Every method targets the same statistic: for each `idu` stratum, the risk
$\theta_g = 1 - \hat S_g(\tau)$ at $\tau = 10$ years, where $\hat S_g$ is the Kaplan–Meier
survival curve in stratum $g$.

Isolate that in one function used everywhere:

```{r estimand}
# Risk at tau for each stratum (g = 0, 1), returned in order c(IDU=0, IDU=1)
risk_at_t <- function(data, tau = 10) {
  vapply(c(0, 1), function(g) {
    sf <- survfit(Surv(t, delta) ~ 1, data = data[data$idu == g, , drop = FALSE],
                  se.fit = F)
    1 - summary(sf, times = tau, extend = TRUE)$surv
  }, numeric(1))
}

# Risk at tau for each stratum, now weighting to control for confounding
risk_at_t_wt <- function(data, tau = 10) {
  
  ps <- glm(idu ~ age + black + cd4nadir, data=data, family=binomial(link="logit"))$fitted.values
  data$wt <- 1 / (data$idu*ps + (1 - data$idu)*(1 - ps))
  
  vapply(c(0, 1), function(g) {
    sf <- survfit(Surv(t, delta) ~ 1, data = data[data$idu == g, , drop = FALSE], 
                  weights = wt, se.fit = F)
    1 - summary(sf, times = tau, extend = TRUE)$surv
  }, numeric(1))
}

```

We will compare the bootstrap results to the analytic standard error (Greenwood standard error).

In the case of weighting, we will also compare against the jackknife robust standard error
including in the survfit function.

```{r analytic}

# Estimate weights
ps <- glm(idu ~ age + black + cd4nadir, data=dat, family=binomial(link="logit"))$fitted.values
dat$wt <- 1 / (dat$idu*ps + (1 - dat$idu)*(1 - ps))
  summary(dat$wt)
  
# Greenwood: point risk and analytic SE (SE of 1 - S equals SE of S)
greenwood_ref <- function(data, tau = 10) {
  
  out <- sapply(c(0, 1), function(g) {
    sf <- survfit(Surv(t, delta) ~ 1, data = data[data$idu == g, , drop = FALSE], 
                  weights=wt, robust=F)
    s  <- summary(sf, times = tau, extend = TRUE)
    c(risk = 1 - s$surv, se = s$std.err)
  })
  colnames(out) <- c("IDU=0", "IDU=1")
  out
}

gw <- greenwood_ref(dat, tau)
gw

# Jackknife
jackknife_ref <- function(data, tau = 10) {
  out <- sapply(c(0, 1), function(g) {
    sf <- survfit(Surv(t, delta) ~ 1, data = data[data$idu == g, , drop = FALSE], 
                  weights=wt, robust=T)
    s  <- summary(sf, times = tau, extend = TRUE)
    c(risk = 1 - s$surv, se = s$std.err)
  })
  colnames(out) <- c("IDU=0", "IDU=1")
  out
}

jk <- jackknife_ref(dat, tau)
jk

```

Two small helpers used by the resampling schemes below: a safe categorical sampler
(avoids R's `sample()` "single value" trap, that selects a range from 1:value if 
only a single value is supplied), and a routine that reads the jump masses
off a Kaplan–Meier curve — i.e. that turns an estimated survival function into the
discrete distribution you simulate from.

```{r helpers}
# Draw n values from the support `times` with probabilities `prob`
rcat <- function(times, prob, n = 1) {
  times[sample.int(length(times), n, replace = TRUE, prob = prob)]
}

# Discrete distribution implied by a KM curve: P(T = t_j) = S(t_{j-1}) - S(t_j)
km_atoms <- function(time, status) {
  sf   <- survfit(Surv(time, status) ~ 1)
  jump <- c(1, head(sf$surv, -1)) - sf$surv     # 0 at times with no "event" of this type
  data.frame(time = sf$time, prob = jump)
}
```

# Ordinary (case) bootstrap — "Efron's method"

This is the classic nonparametric resampling bootstrap. Treat the $n$ observed 
pairs $(t_i, \delta_i)$ as the population and draw
a new sample of size $n$ *with replacement*, then recompute the risk. 

Efron (1981) showed this is equivalent to independently simulating a fresh event time from $\hat F$ and a fresh censoring time from $\hat G$ for each record and keeping the smaller — so both the event
times and the censoring pattern vary from replicate to replicate. 

It assumes nothing beyond independent, identically distributed subjects. 

This is what both the original `classic` (`boot`) and `case` (`censboot(sim = "ordinary")`) blocks do.

Pseudocode: 

```
for r in 1..R:
    draw indices i_1..i_n uniformly with replacement from 1..n
    D* <- data[i_1..i_n, ]
    theta*_r[g] <- 1 - KM_g(tau) fit on D*      for each stratum g
SE[g] <- sd_r( theta*_r[g] )
```

```{r ordinary}
boot_ordinary <- function(data, R = 1000, tau = 10) {
  n <- nrow(data)
  replicate(R, {
    idx <- sample.int(n, n, replace = TRUE)     # resample whole records
    risk_at_t_wt(data[idx, , drop = FALSE], tau)
  })
}

set.seed(123)
ord_reps <- boot_ordinary(dat, R, tau)
ord_se   <- apply(ord_reps, 1, sd)
ord_se
```

# Conditional bootstrap

**How it works.** The conditional bootstrap holds the *censoring pattern* closer to what
was observed. For each subject it draws a brand-new event time from the Kaplan–Meier
event distribution $\hat F$ (within that subject's stratum), but treats censoring
conditionally: subjects observed to be censored keep their exact censoring time, while
subjects observed to fail are assigned a hypothetical censoring time drawn from the
censoring distribution $\hat G$ restricted to times *after* their observed follow-up. The
new observation is the minimum of the two. Because $\hat F$ may not reach 1 when the
largest time is censored, a probability atom beyond the last observed time absorbs the
residual mass (equivalent to Davison & Hinkley's "add a record past the maximum" device;
`censboot` does this internally — cf. the commented `bind_rows` line in the original).

```
Precompute  F_g <- KM of event times within stratum g       (status = delta)
            G   <- KM of censoring times, pooled             (status = 1 - delta)
for r in 1..R:
    for subject i in stratum g:
        Y0*_i <- draw from F_g                               # new latent event time
        if delta_i == 0:  C*_i <- t_i                        # keep observed censoring
        else:             C*_i <- draw from G on (t_i, inf)  # hypothetical censoring
        t*_i     <- min(Y0*_i, C*_i)
        delta*_i <- 1{ Y0*_i <= C*_i }
    theta*_r[g] <- 1 - KM_g(tau) fit on (t*, delta*)
SE[g] <- sd_r( theta*_r[g] )
```

```{r conditional}
boot_conditional <- function(data, R = 1000, tau = 10) {
  n <- nrow(data)
  # F-hat: event-time distribution within each stratum
  Fhat <- lapply(c(0, 1), function(g) {
    d <- data[data$idu == g, ]
    km_atoms(d$t, d$delta)
  })
  names(Fhat) <- c("0", "1")
  # G-hat: censoring-time distribution, pooled across strata
  Ghat      <- km_atoms(data$t, 1 - data$delta)
  Ghat      <- Ghat[Ghat$prob > 0, ]            # keep only the censoring atoms
  tail_time <- max(data$t) + 1                  # absorbs residual mass of F-hat

  one_rep <- function() {
    # Step 1: new event times from the stratum-specific F-hat
    y0 <- numeric(n)
    for (g in c(0, 1)) {
      idx <- which(data$idu == g)
      A   <- Fhat[[as.character(g)]]
      y0[idx] <- rcat(c(A$time, tail_time), c(A$prob, 1 - sum(A$prob)), length(idx))
    }
    # Step 2: censoring times, conditional on the observed pattern
    cstar <- data$t                             # delta == 0: keep observed censoring time
    for (i in which(data$delta == 1)) {         # delta == 1: draw C > observed t_i
      elig     <- Ghat[Ghat$time > data$t[i], ]
      cstar[i] <- if (nrow(elig) == 0) Inf else rcat(elig$time, elig$prob, 1)
    }
    # Step 3: combine and evaluate the estimand
    risk_at_t_wt(data.frame(t = pmin(y0, cstar),
                            delta = as.numeric(y0 <= cstar),
                            idu = data$idu,
                            age = data$age,
                            black = data$black,
                            cd4nadir = data$cd4nadir), tau)
  }
  replicate(R, one_rep())
}

set.seed(123)
cond_reps <- boot_conditional(dat, R, tau)
cond_se   <- apply(cond_reps, 1, sd)
cond_se
```

# Weird bootstrap

**How it works.** The weird bootstrap (Andersen, Borgan, Gill & Keiding 1993) works on the
counting-process representation and holds *everything* fixed except the number of events.
At each observed failure time $u_j$ the risk set $n_j$ and the observed number of events
$d_j$ are known; it redraws the events there as an independent
$\text{Binomial}(n_j, d_j/n_j)$ and rebuilds the survival curve as the product-limit
$\hat S^*(\tau) = \prod_{u_j \le \tau}(1 - d_j^*/n_j)$. Nothing about who is at risk or when
censoring happens changes — only the event counts fluctuate — so it reproduces the
Nelson–Aalen / Greenwood variance by construction. It is "weird" because a replicate
corresponds to no actual set of individuals.

```
Precompute per stratum g:  failure times u_1<...<u_m (<= tau),
                           risk sets n_j, event counts d_j
for r in 1..R:
    for each stratum g, each failure time u_j:
        d*_j <- Binomial(n_j, d_j / n_j)        # risk set n_j held fixed
    theta*_r[g] <- 1 - prod_j (1 - d*_j / n_j)
SE[g] <- sd_r( theta*_r[g] )
```

```{r weird}
boot_weird <- function(data, R = 1000, tau = 10) {
  # Per stratum, capture the fixed risk sets and event counts up to tau
  strata <- lapply(c(0, 1), function(g) {
    sf <- survfit(Surv(t, delta) ~ 1, data = data[data$idu == g, ])
    keep <- sf$n.event > 0 & sf$time <= tau
    list(n = sf$n.risk[keep], d = sf$n.event[keep])
  })
  replicate(R, vapply(strata, function(s) {
    dstar <- rbinom(length(s$d), size = s$n, prob = s$d / s$n)   # resample events only
    1 - prod(1 - dstar / s$n)                                    # 1 - S*(tau)
  }, numeric(1)))
}

set.seed(123)
weird_reps <- boot_weird(dat, R, tau)
weird_se   <- apply(weird_reps, 1, sd)
weird_se

# Now with weights
boot_weird_wt <- function(data, R = 1000, tau = 10) {
  # Per stratum, capture the fixed risk sets and event counts up to tau
  strata <- lapply(c(0, 1), function(g) {
    # It seems to me that you need to apply the weights here because there are not individuals
    #   after resampling. BUT then the n and d are no loner integers. How off will results be if we round(n)?
    #   Results are still closer to the Greenwood (wrong) SE than the other methods
    ps <- glm(idu ~ age + black + cd4nadir, data=data, family=binomial(link="logit"))$fitted.values
    data$wt <- 1 / (data$idu*ps + (1 - data$idu)*(1 - ps))
    
    sf <- survfit(Surv(t, delta) ~ 1, data = data[data$idu == g, ], weights=wt, se.fit=F)
    keep <- sf$n.event > 0 & sf$time <= tau
    list(n = round(sf$n.risk[keep]), d = sf$n.event[keep])
  })
  replicate(R, vapply(strata, function(s) {
    dstar <- rbinom(length(s$d), size = s$n, prob = s$d / s$n)   # resample events only
    1 - prod(1 - dstar / s$n)                                    # 1 - S*(tau)
  }, numeric(1)))
}

set.seed(123)
weird_reps <- boot_weird_wt(dat, R, tau)
weird_se   <- apply(weird_reps, 1, sd)
weird_se


```

# Results: all four standard errors side by side

```{r results}
comparison <- data.frame(
  Stratum          = c("IDU = 0", "IDU = 1"),
  `Risk at 10y`    = round(gw["risk", ], 3),
  `SE: Greenwood`  = round(gw["se", ],   3),
  `SE: Jackknife`  = round(jk["se", ],   3),
  `SE: Ordinary`   = round(ord_se,       3),
  `SE: Conditional`= round(cond_se,      3),
  `SE: Weird`      = round(weird_se,     3),
  check.names = FALSE, row.names = NULL
)
knitr::kable(comparison, caption = "10-year risk of AIDS/death and its standard error.")
```

The weird bootstrap should sit essentially on top of Greenwood; the ordinary and
conditional schemes approximate it up to Monte-Carlo error. Where the conditional and
ordinary SEs *diverge* from Greenwood is the substantive question the paper is after —
e.g. whether they carry variability (such as in estimated weights) that the analytic
formula and the weird scheme do not.


---

*Conceptual background: the vault Main Note "Bootstrapping Censored Data". Full annotated
bibliography: `_corpus/censor-bootstrap_references.md`.*
